---
title: "nyc-temperature-forecasting-analysis"
author: "Alex Heck, Tim McDonald, Cooper Foster"
date: "2025-02-27"
output: html_document
---

Reading in the data and setting up the initial ts object:

```{r}
library(dplyr)
library(forecast)
library(ggplot2)

data <- read.csv('data/updated_set.csv')

data$date <- as.Date(data$date)

#Leap days removed

d1 <- data %>%
  filter(date != "2020-02-29" & date != "2024-02-29")

high.ts <- ts(d1$high, start = c(2020, 1), end = c(2025, 53), frequency = 365)
  

```

For all models below, we will validate by forecasting one day at a time, and then adding the previous day's observed value to the training set as we forecast the next day, and so on. This is designed the replicate the process of actually deploying this model: forecasting just one day ahead, and including today's data for tomorrow's forecast.

# S-Naive Forecast

```{r}
forecast_dates <- seq(as.Date("2025-01-22"), as.Date("2025-02-21"), by = "day")

# Vectors to store forecasts and actual observed values
forecasts <- numeric(length(forecast_dates))
actuals   <- numeric(length(forecast_dates))

# Loop over each forecast date
for (i in seq_along(forecast_dates)) {
  current_date <- forecast_dates[i]
  
  # Training data: all observations up to the day before the forecast date
  train_end_date <- current_date - 1
  
  # Convert the training end date to ts time format (decimal year)
  year_val <- as.numeric(format(train_end_date, "%Y"))
  day_of_year <- as.POSIXlt(train_end_date)$yday + 1
  ts_end <- year_val + (day_of_year - 1) / 365
  
  # Subset the time series using window()
  train_ts <- window(high.ts, end = ts_end)
  
  # Fit a seasonal naive model and generate a 1-day ahead forecast
  forecast_result <- snaive(train_ts, h = 1)
  forecasts[i] <- as.numeric(forecast_result$mean)
  
  # Convert the forecast date into ts time format to extract the actual value
  forecast_year <- as.numeric(format(current_date, "%Y"))
  forecast_day <- as.POSIXlt(current_date)$yday + 1
  ts_forecast_time <- forecast_year + (forecast_day - 1) / 365
  
  # Extract the actual observed value from high.ts for the forecast date
  actual_value <- window(high.ts, start = ts_forecast_time, end = ts_forecast_time)
  actuals[i] <- as.numeric(actual_value)
}

# Combine forecasts and actuals into a data frame for comparison
snaive.res <- data.frame(date = forecast_dates,
                      forecast = forecasts,
                      actual = actuals)

accuracy(snaive.res$forecast, snaive.res$actual)


```

# Linear Regression Model

```{r}

forecasts <- numeric(length(forecast_dates))
actuals   <- numeric(length(forecast_dates))

for (i in seq_along(forecast_dates)) {
  current_date <- forecast_dates[i]
  
  train_end_date <- current_date - 1
  
  year_val <- as.numeric(format(train_end_date, "%Y"))
  day_of_year <- as.POSIXlt(train_end_date)$yday + 1
  ts_end <- year_val + (day_of_year - 1) / 365
  
  train_ts <- window(high.ts, end = ts_end)
  
  model_fit <- tslm(train_ts ~ season)

  forecast_result <- forecast(model_fit, h = 1)
  forecasts[i] <- as.numeric(forecast_result$mean)

  forecast_year <- as.numeric(format(current_date, "%Y"))
  forecast_day <- as.POSIXlt(current_date)$yday + 1
  ts_forecast_time <- forecast_year + (forecast_day - 1) / 365

  actual_value <- window(high.ts, start = ts_forecast_time, end = ts_forecast_time)
  actuals[i] <- as.numeric(actual_value)
}

lm.res <- data.frame(date = forecast_dates,
                      forecast = forecasts,
                      actual = actuals)

accuracy(lm.res$forecast, lm.res$actual)



```

# ARIMA Tuning

To select ideal parameters for ARIMA model, we ran through several validation periods in 2023 and 2024 using the loop below.

```{r}
n_obs <- length(high.ts)
dates <- as.Date("2020-01-01") + 0:(n_obs - 1)

errors.df <- data.frame(Date = as.Date(character()),
                        ME = numeric(),
                        RMSE = numeric(),
                        MAE = numeric(),
                        MPE = numeric(),
                        MAPE = numeric(),
                        stringsAsFactors = FALSE)

nValid <- 2

for (i in seq_along(dates)) {
  current_date <- dates[i]
  current_year <- as.numeric(format(current_date, "%Y"))
  
  if (current_year %in% c(2023, 2024)) {
    if (i + nValid - 1 <= n_obs) {
      
      train.ts <- high.ts[dates < current_date]

      valid.ts <- high.ts[dates %in% dates[i:(i + nValid - 1)]]

      if (length(train.ts) > 0) {
        model <- auto.arima(train.ts, seasonal = TRUE)

        fc <- forecast(model, h = nValid)

        acc <- accuracy(fc, valid.ts)
        me   <- acc["Test set", "ME"]
        rmse <- acc["Test set", "RMSE"]
        mae  <- acc["Test set", "MAE"]
        mpe  <- acc["Test set", "MPE"]
        mape <- acc["Test set", "MAPE"]

        errors.df <- rbind(errors.df,
                           data.frame(Date = current_date,
                                      ME = me,
                                      RMSE = rmse,
                                      MAE = mae,
                                      MPE = mpe,
                                      MAPE = mape))
      }
    }
  }
}

autoplot(fc)
```

# ARIMA Model

```{r}
forecasts <- numeric(length(forecast_dates))
actuals   <- numeric(length(forecast_dates))

for (i in seq_along(forecast_dates)) {
  current_date <- forecast_dates[i]

  train_end_date <- current_date - 1

  year_val <- as.numeric(format(train_end_date, "%Y"))
  day_of_year <- as.POSIXlt(train_end_date)$yday + 1
  ts_end <- year_val + (day_of_year - 1) / 365

  train_ts <- window(high.ts, end = ts_end)

  model_fit <- Arima(train_ts, order = c(5, 0, 0))

  forecast_result <- forecast(model_fit, h = 1)
  forecasts[i] <- as.numeric(forecast_result$mean)

  forecast_year <- as.numeric(format(current_date, "%Y"))
  forecast_day <- as.POSIXlt(current_date)$yday + 1
  ts_forecast_time <- forecast_year + (forecast_day - 1) / 365

  actual_value <- window(high.ts, start = ts_forecast_time, end = ts_forecast_time)
  actuals[i] <- as.numeric(actual_value)
}

arima.res <- data.frame(date = forecast_dates,
                      forecast = forecasts,
                      actual = actuals)

accuracy(arima.res$forecast, arima.res$actual)


```

# NeuralNet Tuning

Similar to the ARIMA model, we tuned by using multiple validation periods to collect the best parameters. However, due to the `nnetar()` function being more computationally intensive than `arima()`, we ran far fewer validations. The chunk below has been purposefully disabled as it takes a long time to run through the parameter ranges.

```{r, eval=FALSE}
p_vals <- 5:15
P_vals <- 0:2
size_vals <- c(5, 7, 9)

best_mape <- Inf
best_params <- list()

nValid <- 2

all_rolling_indices <- which(format(dates, "%Y") %in% c("2023", "2024"))
all_rolling_indices <- all_rolling_indices[all_rolling_indices + nValid - 1 <= length(high.ts)]

if (length(all_rolling_indices) > 7) {
  rolling_indices <- all_rolling_indices[round(seq(1, length(all_rolling_indices), length.out = 7))]
} else {
  rolling_indices <- all_rolling_indices
}

for (p_candidate in p_vals) {
  for (P_candidate in P_vals) {
    for (size_candidate in size_vals) {

      mape_vec <- c()

      for (i in rolling_indices) {
        current_date <- dates[i]
        
        train_ts_cv <- high.ts[dates < current_date]
        valid_ts_cv <- high.ts[dates %in% dates[i:(i + nValid - 1)]]

        if (length(train_ts_cv) > 0) {
          model_candidate <- nnetar(train_ts_cv, repeats = 20, p = p_candidate, P = P_candidate, size = size_candidate)
          fc_candidate <- forecast(model_candidate, h = nValid)

          acc <- accuracy(fc_candidate, valid_ts_cv)
          current_mape <- acc["Test set", "MAPE"]

          mape_vec <- c(mape_vec, current_mape)
        }
      }

      if (length(mape_vec) > 0) {
        avg_mape <- mean(mape_vec, na.rm = TRUE)

        if (avg_mape < best_mape) {
          best_mape <- avg_mape
          best_params <- list(p = p_candidate, P = P_candidate, size = size_candidate)
        }
      }
      
    }
  } 
} 

print(best_params)


```

# NeuralNet Model

```{r}
forecasts <- numeric(length(forecast_dates))
actuals   <- numeric(length(forecast_dates))

for (i in seq_along(forecast_dates)) {
  current_date <- forecast_dates[i]

  train_end_date <- current_date - 1

  year_val <- as.numeric(format(train_end_date, "%Y"))
  day_of_year <- as.POSIXlt(train_end_date)$yday + 1
  ts_end <- year_val + (day_of_year - 1) / 365

  train_ts <- window(high.ts, end = ts_end)

  model_fit <- nnetar(train_ts, p = 14, P = 2, size = 5, repeats = 20)

  forecast_result <- forecast(model_fit, h = 1)
  forecasts[i] <- as.numeric(forecast_result$mean)

  forecast_year <- as.numeric(format(current_date, "%Y"))
  forecast_day <- as.POSIXlt(current_date)$yday + 1
  ts_forecast_time <- forecast_year + (forecast_day - 1) / 365

  actual_value <- window(high.ts, start = ts_forecast_time, end = ts_forecast_time)
  actuals[i] <- as.numeric(actual_value)
}

nn.res <- data.frame(date = forecast_dates,
                      forecast = forecasts,
                      actual = actuals)

accuracy(nn.res$forecast, nn.res$actual)
```

# Naive Forecast

```{r}
forecast_dates <- seq(as.Date("2025-01-22"), as.Date("2025-02-21"), by = "day")

forecasts <- numeric(length(forecast_dates))
actuals   <- numeric(length(forecast_dates))

for (i in seq_along(forecast_dates)) {
  current_date <- forecast_dates[i]

  train_end_date <- current_date - 1

  year_val <- as.numeric(format(train_end_date, "%Y"))
  day_of_year <- as.POSIXlt(train_end_date)$yday + 1
  ts_end <- year_val + (day_of_year - 1) / 365

  train_ts <- window(high.ts, end = ts_end)

  naive_fit <- naive(train_ts, h = 1)
  forecasts[i] <- as.numeric(naive_fit$mean)

  forecast_year <- as.numeric(format(current_date, "%Y"))
  forecast_day <- as.POSIXlt(current_date)$yday + 1
  ts_forecast_time <- forecast_year + (forecast_day - 1) / 365

  actual_value <- window(high.ts, start = ts_forecast_time, end = ts_forecast_time)
  actuals[i] <- as.numeric(actual_value)
}

naive.res <- data.frame(date = forecast_dates,
                      forecast = forecasts,
                      actual = actuals)

accuracy(naive.res$forecast, naive.res$actual)

```

